Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-08-23 analysis/paper/paper-doc.qmd
move all 002-coefficient-variation code into the qmd and delete that file

---
title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
author:
  - name: Gayoung Park
    email: gayoungp@uw.edu
    affiliations: University of Washington
  - name: Ben Marwick
    email: bmarwick@uw.edu
    affiliations: University of Washington
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
editor: visual
date: now
date-format: long
title-block-published: "Last updated" 
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

@@

1.  Distribution of attributes

@@

2.  Coefficient of variation

We used CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from ` min_cv` to ` max_cv` (@fig-four). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes (@fig-three), the CV of body length and maximum length have the lowest values. The lower CV of maximum length indicates highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes on the artifacts. Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

![The points and blue colored numbers represent CV values for each attribute measured on the stemmed points. 95% confidence intervals for the CVs were computed using Sharma & Krishna's method. The vertical lines indicate the range of confidence intervals.](images/003-cv-sharma-01.png){#fig-four width="339"}

3.  Principal Component Analysis

@@

After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-six). Body length of Phase 2, maximum length, mid width, and tang width are under 25, which indicates as a result of indirect bias, while only maximum length are under 25 in Figure 4. On the other hand, Body length of Phase 3, stem length, stem width, and tang length are relatively higher then 25 and mostly over 30 indicating guided variation.

@@

![CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.](images/004-cv-phases-01.png){#fig-six width="346"}

@@

We examined the CV values per site to address our research question about the cultural transmission over space. We excluded assemblages that contain a single point. @fig-nine shows that Sachang site has the highest CV values with the widest confidence level. Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2. We excluded sites that have only one stemmed point for the analysis. Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points. CV values of body length is the lowest while stem width the highest among all assemblages. Overall, there are no clear patterns between the sites.

![CV values per site that has more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.](images/004-cv-sites-01.png){#fig-nine width="519"}

>>>>

---
title: "Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea"
author:
  - Gayoung Park:
      institute: uw
      correspondence: "yes"
      email: gayoungp@uw.edu
  - Ben Marwick:
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: Seattle, WA, USA
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
editor: visual
date: now
date-format: long
title-block-published: "Last updated" 
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

@@

## Distribution of attributes

@@

## Coefficient of variation

```{r}
#| label: prepare-for-cv-calcs

# Coefficient of variation (CV): the ratio of the standard deviation divided by mean

cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100

#try corrected cv with different methods

library(cvcqv)
ccv <- function(x, ... ) cv_versatile(x, method = "kelly", ...)
mccv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100 * (1 + 1/(4*length(x)))

# we've got DF a data frame with cols as variables
# rows and specimens, and values as dimensions
# need to add a column of site ID

DFCV <- map_dbl(DF, cv)

#Same as map but better view for summary
cv_all <- map_df(DF, cv)
#arrange the data properly
cv_data_all <- gather(cv_all)

# to install MKmisc
## Install package BiocManager
# install.packages("BiocManager")
## Use BiocManager to install limma
# BiocManager::install("limma")
# install.packages("MKmisc")

library(MKmisc)

# CV value
sharma_cv <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$estimate*100

#CV sharma interval values
sharma_int_low <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[1]*100
sharma_int_high <- function(x,...) cvCI(x, conf.level = 0.95, method = "sharma", na.rm = FALSE)$conf.int[2]*100

all_cv_sharma <- map_dbl(DF, sharma_cv)
all_low_sharma <- map_dbl(DF, sharma_int_low)
all_high_sharma <- map_dbl(DF, sharma_int_high)

#create data frame
cv_data_sharma <- data.frame(all_cv_sharma, all_low_sharma, all_high_sharma)

#add column name
library(tibble)
cv_data_sharma_named <- rownames_to_column(cv_data_sharma, var = "attribute")

# for in-text numbers
max_cv <- round(max(cv_data_sharma_named$all_cv_sharma), 1)
min_cv <- round(min(cv_data_sharma_named$all_cv_sharma), 1)

```

We used CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes (@fig-plot-attributes), the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

```{r}
#| label: fig-plot-cv-all-attributes
#| fig-cap: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points. 95% confidence intervals for the CVs were computed using Sharma & Krishna's method. The vertical lines indicate the range of confidence intervals.
#| fig-width:   4.8
#| fig-height:  4.5

ggplot(cv_data_sharma_named,
       aes(attribute,
           all_cv_sharma)) +        # ggplot2 plot with confidence intervals
  geom_point() +
  geom_errorbar(aes(ymin = all_low_sharma, 
                    ymax = all_high_sharma), 
                width = 0.2) +
  ylab("CV") +
  xlab("Variables") +
  geom_text(aes(label = round(all_cv_sharma,1)), 
            col="blue", 
            hjust = -0.3, 
            size = 3) +
  theme_bw()
```

## Principal Component Analysis

@@

After excluding the Phase 1 stemmed points, overall CV values become lower (@fig-plot-cv-over-time). Body length of Phase 2, maximum length, mid width, and tang width are under 25, which indicates as a result of indirect bias, while only maximum length are under 25 in Figure 4. On the other hand, Body length of Phase 3, stem length, stem width, and tang length are relatively higher then 25 and mostly over 30 indicating guided variation.

@@

```{r}
#| label: tab-cv-over-time

#-- temporal patterns ----------------------------------------------------------------
# Make a table of CVs for all variable grouped by stage, according to
# stemmed point chronology
# Exclude the first phase that has only two artefacts

cv_by_stage_df  <-
  df_sitename %>%
  dplyr::select(-sitename,
         -SPstage.Raw_material) %>%
  group_by(SPstage.Stage) %>%
  add_tally() %>%
  dplyr::filter(SPstage.Stage > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Stage,
               names_to = "variable",
               values_to = "value") %>%
  group_by(SPstage.Stage, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

####### significant test between the two phases ######
library(cvequality)

# created nested tibble, nest by variable
df_sitename_nested <-
  df_sitename %>%
  dplyr::select(-sitename, -SPstage.Raw_material) %>%
  group_by(SPstage.Stage) %>%
  add_tally() %>%
  dplyr::filter(SPstage.Stage > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Stage,
               names_to = "variable",
               values_to = "value") %>%
  ungroup() %>%
  group_by(variable) %>%
  nest()

# apply mslr test to each variable (row with a tibble)
# Krishnamoorthy and Lee’s (2014) modified signed-likelihood ratio test
# this is a table we can show in the text
table_mslr_test_results_for_phase_differences <-
  df_sitename_nested %>%
  mutate(mslr_result = map_df(data, ~mslr_test(nr = 1e4,
                                               .$value,
                                               .$SPstage.Stage))) %>%
  unnest(mslr_result) %>%
  dplyr::select(-data)

names(table_mslr_test_results_for_phase_differences) <- 
  c("Variable", "MSLR statistic", "p value")

knitr::kable(table_mslr_test_results_for_phase_differences,
             caption = "CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.")
```

```{r}
#| label: fig-plot-cv-over-time
#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   4.5
#| fig-height:  4.5

# facet plot for two phases
cv_by_stage_df %>%
  ggplot(aes(x = factor(SPstage.Stage),
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Phase") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ variable, scales = "free_y") +
  theme_bw()
```

@@

We examined the CV values per site to address our research question about the cultural transmission over space. We excluded assemblages that contain a single point. @fig-plot-cv-by-site shows that Sachang site has the highest CV values with the widest confidence level. Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2. We excluded sites that have only one stemmed point for the analysis. Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points. CV values of body length is the lowest while stem width the highest among all assemblages. Overall, there are no clear patterns between the sites.

```{r}
#| label: fig-plot-cv-by-site
#| fig-cap: CV values per site that has more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   8.5
#| fig-height:  4.5

cv_by_site_df  <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, 
                -lat_dd, 
                -long_dd, 
                -sitename, 
                -SPstage.Raw_material) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-full_sitename,
               names_to = "variable",
               values_to = "value") %>%
  group_by(full_sitename, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

## CV for per each site (n=X) with the full site name : label

cv_by_full_site_df_label <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  filter( n > 1) %>%
  mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
  dplyr::select(-n) %>%
  nest(-full_sitename, -label) %>%
  mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>%
  dplyr::select (-cv_by_site)

## add label to main dataframe

cv_plot_site_label <- cv_by_site_df %>%
  left_join (cv_by_full_site_df_label) %>%
  dplyr::select (- data,-full_sitename)

## facet plot

cv_plot_site_label %>%
  ggplot(aes(x = variable,
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Attribute") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 6.5))
```
